Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-07-30 manuscript/manuscript.Rmd
trying to fill in discussion and case reports...
but it's not very good

@@

all |> ungroup() |>  left_join(names) |> select(Paper=name_pretty, `All reps`=orig_v_other, "1st rep and Rescue"=orig_v_rr, "Rescue" = orig_v_rescue, "Non-rescue"=orig_v_not_rescue, "Other reps"=rescue_v_reps) |> knitr::kable(format="latex", booktabs=T, linesep="", caption="$P$-original values between different sets of experiments. The primary analysis is between the original result and the meta-analytic aggregation of all replications. All $p$-original values assume an imputed heterogeneity value of $\\tau$=.21.", digits=3) |> 
  add_header_above(c(" "=1, "Original and"=4, "Rescue and"=1)) |> add_header_above(c(" "=1, "p-original comparing between"=5))

@@

As secondary measures, we calculated the $p$-original values between a) the original and just the first replication and rescue, b) the original and the rescue, c) the original and non-rescue replications, and d) the rescue and other replications.
For a) the original versus the rescue, the median value of $p$-original was `r rr_summ$median |> round(2)` [IQR: `r rr_summ$low |> round(2)` - `r rr_summ$high|> round(2)`], and `r round(rr_summ$less_than*100)`% of the $p$-original values were less than .05.
For b) the original versus the rescue, the median value of $p$-original was `r rescue_summ$median |> round(2)` [IQR: `r rescue_summ$low |> round(2)` - `r rescue_summ$high|> round(2)`], and `r round(rescue_summ$less_than*100)`% of the $p$-original values were less than .05.
For c) the original versus all the non-rescue replications, the median value of $p$-original was `r rep_summ$median |> round(2)` [IQR: `r rep_summ$low |> round(2)` - `r rep_summ$high |> round(2)`], and `r round(rep_summ$less_than*100)`% of the $p$-original values were less than .05.
For d) the rescue versus the other replications, the median value of $p$-original was `r rescue_rep_summ$median |> round(2)` [IQR: `r rescue_rep_summ$low |> round(2)` - `r rescue_rep_summ$high |> round(2)`], and `r round(rescue_rep_summ$less_than*100)`% of the $p$-original values were less than .05.

@@

Given the mix of successful and unsuccessful rescue projects, we discuss a few projects where we have speculations about why they turned out the way they did.

@@

@ngo2019 was a second successful rescue. The original study found a large effect; hence the first replication, powering for 80% power on the reported effect, recruited a small sample of 12 people. This small study failed to find the original effect.
The rescue, powered using 2.5x the original sample [as recommended by @simonsohn2015], recovered a clear effect (albeit a much smaller one).
There are reasons to think that some effect sizes in the literature may be inflated [@opensciencecollaboration; @camerer2018], and it is also possible that changes to experiments or switches to online could result in noisier samples and thus smaller effect sizes.
Therefore, replications with smaller samples than the original (even if powered to the original effect size), may not be very diagnostic, and could potentially benefit from a re-replication.

Not all rescues of small replications succeeded, however.
@payne2008 was a study of the effects of sleep on memory consolidation. They  showed participants a number of images and then hours later (after either sleep or no sleep) measured their recall for parts of the images.
The first replication struggled to recruit participants to complete a followup session after sleep and only reported a sample size of 23 (the original had 48).
The rescue attempted to recruit a larger sample (target 88), but also experienced difficulties getting participants to complete the second part of the experiment 12 hours after the first, managing also only to recruit 23 people.
The lesson here may be that sleep research is difficult to conduct online.
However, an online replication by @denis2022 reports qualitatively similar (but quantitatively smaller) results to @payne2008 on related but not identical measures.

@@

@@

The rescues all showed smaller effect sizes than their original studies, regardless of whether the pattern of effects replicated.
A large minority of replications had effect sizes that were statistically inconsistent with the original effect, even accounting for expected levels of heterogeneity between studies.
These diminished and inconsistent effects suggest that even if a re-replication "works", it may be difficult to build upon as follow-up studies will need large samples to detect small effects.

@@

The reported rescue projects are a small sample of replications, leading to highly uncertain estimates. 
Additionally, they are also chosen non-randomly, as they have been selected for twice based on student interest -- once for the original replication and again for the rescue.
However, this selection bias is likely to correlate with how graduate students and other early career researchers choose what topics to work on and what studies to build on.
Nevertheless, our estimate is still broadly consistent with that of ManyLabs 5 [29% vs. 22%, @ebersole2020].

@@

Due to the limited number of replications that are conducted and systematically reported, and the even smaller number of re-replications reported, there is a lot of uncertainty in how much replicability varies across features of the original experiment and replications. 
## Conclusion
<!-- were false positives, we merely claim the original results do not support cumulative research by early-career researchers under the constrained conditions we tested. -->

>>>>

We first discuss the overall rate of replication in the rescue projects, followed by looking at the relative effect sizes of the rescues and correlates of rescue success. In the last subsection we describe the topic of each studies and focus on a few interesting case studies.

@@

all |> ungroup() |>  left_join(names) |> select(Paper=name_pretty, `All reps`=orig_v_other, "1st Rep and Rescue"=orig_v_rr, "Rescue" = orig_v_rescue, "Non-rescue"=orig_v_not_rescue, "Other reps"=rescue_v_reps) |> knitr::kable(format="latex", booktabs=T, linesep="", caption="$P$-original values between different sets of experiments. The primary analysis is between the original result and the meta-analytic aggregation of all replications. All $p$-original values assume an imputed heterogeneity value of $\\tau$=.21.", digits=3) |> 
  add_header_above(c(" "=1, "Original and"=4, "Rescue and"=1)) |> add_header_above(c(" "=1, "p-original comparing between"=5))

@@

As secondary measures, we calculated the $p$-original values between a) the original and just the first replication and rescue, b) the original and the rescue, c) the original and non-rescue replications, and d) the rescue and other replications.
For a) the original versus the first replication and the rescue, the median value of $p$-original was `r rr_summ$median |> round(2)` [IQR: `r rr_summ$low |> round(2)` - `r rr_summ$high|> round(2)`], and `r round(rr_summ$less_than*100)`% of the $p$-original values were less than .05.
For b) the original versus the rescue, the median value of $p$-original was `r rescue_summ$median |> round(2)` [IQR: `r rescue_summ$low |> round(2)` - `r rescue_summ$high|> round(2)`], and `r round(rescue_summ$less_than*100)`% of the $p$-original values were less than .05.
For c) the original versus all the non-rescue replications, the median value of $p$-original was `r rep_summ$median |> round(2)` [IQR: `r rep_summ$low |> round(2)` - `r rep_summ$high |> round(2)`], and `r round(rep_summ$less_than*100)`% of the $p$-original values were less than .05.
For d) the rescue versus the other replications, the median value of $p$-original was `r rescue_rep_summ$median |> round(2)` [IQR: `r rescue_rep_summ$low |> round(2)` - `r rescue_rep_summ$high |> round(2)`], and `r round(rescue_rep_summ$less_than*100)`% of the $p$-original values were less than .05.

@@

Given the mix of successful and unsuccessful rescue projects, we briefly discuss all the projects and go more in depth on a few projects where we have speculations about why they turned out the way they did.

@@

@ngo2019 was a second successful rescue. The original study found a large effect; hence the first replication, powering for 80% power on the reported effect, recruited a small sample of 12 people. This small study failed to find the original effect.
The rescue, powered using 2.5x the original sample [as recommended by @simonsohn2015], recovered a clear effect (albeit a much smaller one). An additional replication from the literature, @ngo2021 successfully replicated @ngo2019. 
There are reasons to think that some effect sizes in the literature may be inflated [@opensciencecollaboration; @camerer2018], and it is also possible that changes to experiments or switches to online could result in noisier samples and thus smaller effect sizes.
Therefore, replications with smaller samples than the original (even if powered to the original effect size), may not be very diagnostic, and could potentially benefit from a re-replication.

Not all rescues of small replications succeeded, however.
@payne2008 was a study of the effects of sleep on memory consolidation. They  showed participants a number of images and then hours later (after either sleep or no sleep) measured their recall for parts of the images.
The first replication struggled to recruit participants to complete a followup session after sleep and only reported a sample size of 23 (the original had 48).
The rescue attempted to recruit a larger sample (target 88), but also experienced difficulties getting participants to complete the second part of the experiment 12 hours after the first, managing also only to recruit 23 people.
The lesson here may be that sleep research is difficult to conduct online.
However, an online replication by @denis2022 reports qualitatively similar (but quantitatively smaller) results to @payne2008 on related but not identical measures. @denis2022 which has an overlapping author team to @payne2008 is included here as an additional replication.

@@

@birch2007 is about a false belief task comparing ignorance and plausible knowledge conditions. The first replication, an additional class replication, and the rescue all found effects directionally consistent with the original, but much smaller.

@craig2014 had White Americans read an article about shifting racial demographics in the US, either with or without a framing that assuaged the potential threat, and found that reading the un-assuaged version led to the endorsement of more conservative positions.

@gong2019 looked at the interplay between age and willingness to donate in older adults in Hong Kong and found a three-way interaction between age, type of donation, and relationship to recipient. Neither the replication that sampled US adults nor the rescue that sampled East Asian adults recovered this interaction.

@haimovitz2016 manipulated parents' mindsets with "failure-is-enhancing" or "failure-is-debilitating" primes and looked at their reactions to a hypothetical situation in which a child received a failing grade. The first replication was directionally consistent with the original, but an additional class replication was not. The rescue added manipulation checks which passed, but did not recover the target effect.

@hopkins2016 examined whether people prefer reductive explanations to horizontal explanations. While the original found a preference for reductive explanations, the replication and rescue both did not.

@jara-ettinger2022 looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape over color or material. The replication and rescue both found an above chance preference for shape, but with widely different item-level effects and relative preferences for color and material.

@paxton2012 found that participants who completed the CRT before the moral judgement task gave more utilitarian answers than those who did the moral judgment task before the CRT. Neither the replication nor the rescue found this effect.

@porter2016 looked at how linguistic intergroup bias influences judgments of shared group membership. The replication was directionally against the original, the rescue found that effect, although the manipulation check failed. 
@schechtman2010 looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more. The first replication and rescue both found the same main effect, but with a very different pattern over distance.

As discussed in methods, @tarampi2016 had participants do a timed navigation task and found an interaction between participant gender and the framing of the task (social or spatial) on performance. Neither first replication nor rescue recovered the interaction. 
@todd2016 looked at participants speed at classifying toys and guns as a function of whether they had just need a White or Black boy's face. @todd2016 found an interaction where Black faces primed faster "gun" responses and White faces faster "toy" responses. When we reached out to the original authors, they offered to run a parallel rescue on their student subject pool, which we have here as an additional replication. The first replication did not find the interaction, but both the better-powered rescue did.

@yeshurun2003 looked at the relationship between spatial attention and temporal resolution. They measured the threshold where two flashes of light are percieved as one flash versus two as a function of whether the location of the flashes was pre-cued or not. @yeshurun2003 found that cueing led to less temporal resolution. This was independently replicated by @chica2009, but was not replicated by either the student replication or the rescue.

@@

The rescues all showed smaller effect sizes than their original studies, regardless of whether the pattern of effects replicated.
A large minority of replications had effect sizes that were statistically inconsistent with the original effect, even accounting for expected levels of heterogeneity between studies.

@@

The reported rescue projects are a small sample of replications, leading to highly uncertain estimates. 
Additionally, they were chosen non-randomly, as they were doubly selected for based on student interest -- once for the original replication and again for the rescue.
However, this selection bias is likely to correlate with how graduate students and other early career researchers choose what topics to work on and what studies to build on.
Nevertheless, our estimate is still broadly consistent with that of ManyLabs 5 [29% vs. 22%, @ebersole2020].

@@

Due to the limited number of replications that are conducted and systematically reported, and the even smaller number of re-replications reported, there is a lot of uncertainty in how much replicability varies across features of the original experiment and replications.  
## Conclusion
Overall, the diminished and inconsistent effects we found in the rescues suggest that even if a re-replication "works", it may be difficult to build upon as follow-up studies will need large samples to detect small effects.

Replications are done for many reasons, and researchers often have some prior beliefs about whether a replication will succeed. These beliefs may influence how a researcher interprets a failed replication and thus whether they think it's worth trying to rescue. For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication and thus want to rescue by tweaking these details. On the other side, a researcher who is dubious of the methods, results, or statistical approaches, or who has experienced failures to replicate and extend the work, may be quicker to interpret a failure as a more definitive negative result. Interpretations may also depend on the perceived strength of the replication and what effect size is cared about.

While we focus here on a narrow type of replication and rescue, that done by an early career researcher with the goal of establishing whether this is a result *they* can build on, it exists as part of a wider ecosystem of replications that are done for different reasons with different properties.
